An AI Success Story in Metaphysics: \\ the Case of the Inconsistency in G\"odel's Ontological Argument

### 2016-01-13 Papers/InProgress/IJCAI/paper.tex
paper improvements

\newcommand{\imp}{\rightarrow}
\newcommand{\biimp}{\leftrightarrow}
\newcommand{\allq}{\forall}
\newcommand{\exq}{\exists}
\newcommand{\seq}{\vdash}

@@

St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz \cite{ToDo:Leibniz}.  G\"odel's manuscript
(Fig.~\ref{GoedelScript}) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5}. The main difference to G\"odel's version is the
addition of a conjunct in the definition of \emph{essence}. G\"odel's
different definition of essence can be seen either in his manuscript
(Fig.~\ref{GoedelScript}) or, in more modern notation, in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required. This difference has been
considered inessential and merely an oversight by many. For instance,
Hazen 1998, p.365 \cite[p.365]{Hazen} states that ``G\"odel left this
clause out in [11], but this appears to have been an oversight--it is
included in related manuscripts''. However, the omitted conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_S5}
where the model finder Nitpick confirms consistency).\footnote{In
  personal communication, Dana Scott confirmed that he was unaware
  that G\"odel's axioms were inconsistent.}

@@

\subsection{Improved Embedding \SFiveU}\label{sec:improvedembedding}
\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/QML_S5.png}}
\caption{Improved Embedding of \SFiveU} \label{QML_S5}
\end{figure}

@@

The new embedding of \SFiveU in Isabelle is shown in Fig.~\ref{QML_S5}.
%
\marginpar{@Bruno: I commented you paragraph here, since I am not
  convinced (yet) that e.g. BF and CBF are equally valid in \SFiveU
  and \SFive}
% \SFiveU is only slightly stronger than \SFive. Most importantly,
% $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi^{U}$, where
% $\varphi^U$ is obtained from $\varphi$ by replacing all $\Box$ and
% $\Dia$ by, respectively, $\BlackBox$ and $\BlackDia$. Therefore,
% \SFiveU can be considered as adequate for metaphysics as \SFive
% (cf. \cite[p. 127]{WilliamsonModalLogicAsMetaphysics}). 
% % \cite{Mattey}, \cite[ToDo]{DunnHardegree}
%
With this improved embedding, the final theorem T3
(\textit{``Necessarily, there exists God''}) can be derived from
Scott's consistent version of the axioms fully automatically.  The fully
automatic proof has been generated (in about 2,5 seconds) by the theorem prover
\textsc{Leo-II}~\cite{C26} and subsequently verified in the proof
assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5}.
With the embedding used by Benzm\"uller and Woltzenlogel-Paleo \shortcite{C40}, the
provers still had to be given the intermediate theorem T2 and the corollary
C in order to manage to prove T3.

@@

Wider adoption of higher-order theorem proving technology for reasoning about and within embedded object logics, especially among non-expert users, is still hindered by the gap between the syntax used by people, when they write logical formulas with pen and paper, and the syntax used by higher-order logic theorem provers. Even when the syntax of the underlying higher-order system is elegant (as is the case in Isabelle/HOL), the embedding of HOML into HOL may easily expose details of HOL that may be uncommon to the user, disturbing his/her experience while using the system. To illustrate this point, Fig.~\ref{UglyEssence} shows how the definition of essence looked like in previous work \cite{J28}, where advanced syntax-sugaring features were not used. It looks notably higher-order, and its style differs significantly from the common style seen in works on modal logics and the ontological argument. The following specific issues can be enumerated:
\begin{enumerate}
\item Lambda abstractions (which are a typical higher-order feature) appear explicitly in places where it didn't need to appear in a pure HOML formulation (cf. G\"odel's manuscript, Fig.~\ref{GoedelScript}).
\item Quantifiers appear as higher-order defined constants, and not as binders. This forces the user to read (and write) formulas of the form $\forall (\lambda x. A(x))$ instead of the more common $\forall x. A(x)$.
\item The lifted modal connectives are represented by prefixing the letter ``m'' (e.g. $m\wedge$ and $m\rightarrow$). The prefix disturbs the user, as it constantly reminds him/her that there is something unusual about the modal connectives.
\item Higher-order parenthesis conventions for the application of a predicate to a term are used. 
Instead of reading $\psi(y)$, as he would expect, he has to read $(\psi y)$. Outside niche areas in computer science, the former syntax is more widely known than the latter.
\end{enumerate}

@@

\subsection{Informal Refutation}

@@

\subsection{Mapping the Inconsistency to G\"odel}
The inconsistency verified in Fig.~\ref{InconsistencyIsabelleK} follows from the definition of
essence ($ess$), the definition of necessary existence ($NE$), the
axioms A1(a) and A2 (which entail theorem T1), and axiom A5. It remains to demonstrate that
these ingredients are actually present in G\"odel's manuscript in
Fig.~\ref{GoedelScript}.

This can easily be seen: Axioms A1(a) in
Fig.~\ref{InconsistencyIsabelleK} is implied by Axiom A2 and the
highlighted footnote remark in Fig.~\ref{GoedelScript}. Axioms A2 and
A5 in Fig.~\ref{InconsistencyIsabelleK} correspond to A4 and A3 in
Fig.~\ref{GoedelScript}. The definitions of essence and necessary
existence are easy to identify. Thus, we conclude that the verified
inconsistency from Fig.~\ref{InconsistencyIsabelleK} does apply to 
G''odel's proof script in Fig.~\ref{GoedelScript}.

\section{Conclusion}\label{sec:conclusion}
The inconsistency of G\"odel's original axioms in his ontological
argument has been detected automatically by the prover
\textsc{Leo-II}. Here we have presented a rational reconstruction and
verification of the inconsistency proof in Isabelle/HOL. This
inconsistency argument applies to all normal modal logics including K.
\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{./Images/Manuscript2.png}}
\caption{G\"{o}del's manuscript, with mutually inconsistent axioms and definitions highlighted.} \label{GoedelScript}
\end{figure*}

Moreover, we have reported on several technical improvements regarding
the semantic embedding approach. In fact, we have a achieved a nearly
perfect match between pen and paper presentations in HOML and the
syntax in Isabelle/HOL. As a result, the embedding of HOML in
HOL is now fully transparent and, as we think, ready for wider adoption by
philosophers.

On the other hand, there is still room for improvements in
Isabelle/HOL. For example, Isabelle/HOL's Sledgehammer tool, which
realises calls to external automated theorem provers, such as
\textsc{Leo-II}, still fails to detect the inconsistency of G\"odel's
axioms, while a direct modeling of the problem in TPTP THF syntax in
combination with a direct call of \textsc{Leo-II} succeeds. In other
words, without independent experiments directly in TPTP THF the
inconsistency would thus not have been detected. There are further
technical issues regarding Sledgehammer that motivate
improvements. For example, in default setting, Sledgehammer does not
inform the user early that a proof has been found and instead silently
applies first a series of time-consuming analysis tools (e.g. it's
minimizer), before it eventually informs the user of success. For the
involved proof problems presented here (e.g. G\"odel's theorem T3:
\textit{Necessarily, there exists God}), this phase of silence may
take several minutes (during which the user might actually give up on
the proof attempt), while \textsc{Leo-II} already reported success to
Sledgehammer much earlier (e.g. after 2,5 seconds for T3).

Moreover, our work points to a challenge for proof transformation:
even if \textsc{Leo-II}'s proof could be automatically verified
in Isabelle, this would clearly not entail that Isabelle/HOL could
also extract from this verification a human-intuitive argument, for
example, an argument as has been presented here. For philosophers,
however, the reconstruction of such human-intuitive arguments from
machine found proofs will likely be crucial for accepting any novel results.

Another open problem that we solved in this paper is a fully automatic
proof of T3 directly from Scott's axioms. Again, this proof was
contributed by \textsc{Leo-II}. However, this has been possible only
after providing a more efficient embedding, based on a universal
accessibility relation, for HOML S5 in HOL.

Both, the automated detection of the inconsistency in G\"odel's axioms
and the fully automatic proof of T3 from Scott's axioms demonstrates
the potential of our AI technology for philosophy: this technology is
already in its current state able to contribute novel results to
metaphysics and to conduct reasoning steps at granularity-levels
beyond common human capabilities.

@@

%% The file named.bst is a bibliography style file for BibTeX 0.99c
\bibliographystyle{named}
\bibliography{Bibliography}

>>>>

\newcommand{\imp}{{\rightarrow}}
\newcommand{\biimp}{\leftrightarrow}
\newcommand{\allq}{\forall}
\newcommand{\exq}{\exists}
\newcommand{\seq}{\vdash}

@@

St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz \cite{ToDo:Leibniz}.  G\"odel's manuscript
(Fig.~\ref{GoedelScript}) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5}. The main difference to G\"odel's version is the
addition of a conjunct in the definition of \emph{essence}. G\"odel's
different definition of essence can be seen either in his manuscript
(Fig.~\ref{GoedelScript}) or, in more modern notation, in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required. This difference has been
considered inessential and merely an oversight by many. For instance,
Hazen 1998, p.365 \cite[p.365]{Hazen} states that ``G\"odel left this
clause out in [11], but this appears to have been an oversight--it is
included in related manuscripts''. However, the omitted conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_S5}
where the model finder Nitpick \cite{Nitpick} confirms consistency).\footnote{In
  personal communication, Dana Scott confirmed that he was unaware
  that G\"odel's axioms were inconsistent.}

@@

\subsection{Improved Embedding}\label{sec:improvedembedding}
\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/QML_S5.png}}
\caption{Improved Embedding of \SFiveU} \label{QML_S5}
\end{figure}

@@

The new embedding of \SFiveU in Isabelle is shown in Fig.~\ref{QML_S5}.
%
\marginpar{@Bruno: I commented you paragraph here, since I am not
  convinced (yet) that e.g. BF and CBF are equally valid in \SFiveU
  and \SFive}
% \SFiveU is only slightly stronger than \SFive. Most importantly,
% $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi^{U}$, where
% $\varphi^U$ is obtained from $\varphi$ by replacing all $\Box$ and
% $\Dia$ by, respectively, $\BlackBox$ and $\BlackDia$. Therefore,
% \SFiveU can be considered as adequate for metaphysics as \SFive
% (cf. \cite[p. 127]{WilliamsonModalLogicAsMetaphysics}). 
% % \cite{Mattey}, \cite[ToDo]{DunnHardegree}
%
With this improved embedding, the final theorem T3
(\textit{``Necessarily, there exists God''}) can be derived from
Scott's consistent version of the axioms fully automatically.  The fully
automatic proof has been generated (in about 2,5 seconds) by the theorem prover
\textsc{Leo-II}~\cite{C26} and subsequently verified in the proof
assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5}.
The collaboration between the two systems has been orchestrated by
Isabelle's Sledgehammer tool \cite{Sledgehammer}.

With the embedding used by Benzm\"uller and Woltzenlogel-Paleo \shortcite{C40}, the
provers still had to be given the intermediate theorem T2 and the corollary
C in order to manage to prove T3.

@@

Wider adoption of higher-order theorem proving technology for reasoning about and within embedded object logics, especially among non-expert users, is still hindered by the gap between the syntax used by people, when they write logical formulas with pen and paper, and the syntax used by higher-order logic theorem provers. Even when the syntax of the underlying higher-order system is elegant (as is the case in Isabelle/HOL), the embedding of HOML into HOL may easily expose details of HOL that may be uncommon to the user, disturbing his/her experience while using the system. To illustrate this point, Fig.~\ref{UglyEssence} shows how the definition of essence looked like in previous work \cite{J28}, where advanced syntax-sugaring features were not used. It looks notably higher-order, and its style differs significantly from the common style seen in works on modal logics and the ontological argument. The following specific issues can be enumerated:
\begin{enumerate}
\item Lambda abstractions (which are a typical higher-order feature) appear explicitly in places where it didn't need to appear in a pure HOML formulation (cf. G\"odel's manuscript, Fig.~\ref{GoedelScript}).
\item Quantifiers appear as higher-order dqefined constants, and not as binders. This forces the user to read (and write) formulas of the form $\forall (\lambda x. A(x))$ instead of the more common $\forall x. A(x)$.
\item The lifted modal connectives are represented by prefixing the
  letter ``m'' (e.g. $m\wedge$ and $m\imp$). The prefix disturbs the
  user, as it constantly reminds him/her that there is something
  unusual about the modal connectives.
\item Higher-order parenthesis conventions for the application of a predicate to a term are used. 
Instead of reading $\psi(y)$, as he would expect, he has to read $(\psi y)$. Outside niche areas in computer science, the former syntax is more widely known than the latter.
\end{enumerate}

@@

\subsection{Informal Argument}

@@

\subsection{Mapping the Inconsistency to G\"odel}
The inconsistency verified in Fig.~\ref{InconsistencyIsabelleK} follows from the definition of
essence (ess), the definition of necessary existence (NE), the
axioms A1(a) and A2 (which entail theorem T1), and axiom A5. It remains to demonstrate that
these ingredients are actually present in G\"odel's manuscript in
Fig.~\ref{GoedelScript}.

This can easily be seen: Axioms A1a in
Fig.~\ref{InconsistencyIsabelleK} is implied by Axiom Ax2 and the
highlighted footnote remark in Fig.~\ref{GoedelScript}. Axioms A2 and
A5 in Fig.~\ref{InconsistencyIsabelleK} correspond to Ax4 and Ax3 in
Fig.~\ref{GoedelScript}. The definitions of essence and necessary
existence are easy to identify. Thus, we conclude that the verified
inconsistency from Fig.~\ref{InconsistencyIsabelleK} does apply to 
G\"odel's original manuscript.

\section{Conclusion}\label{sec:conclusion}
The axioms and definitions in G\"odel's manuscript are inconsistent;
this has been detected automatically by the prover
\textsc{Leo-II}. Here we have presented a rational reconstruction and
verification of the inconsistency argument in Isabelle/HOL. This
argument is valid in all normal HOMLs including base logic K.
\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{./Images/Manuscript2.png}}
\caption{G\"{o}del's manuscript, with mutually inconsistent axioms and definitions highlighted.} \label{GoedelScript}
\end{figure*}

We have also presented several technical improvements regarding the
semantic embedding approach. In particular, we have a achieved a
nearly perfect match between pen and paper presentations in HOML and
the syntax in Isabelle/HOL. As a result, the embedding of HOML in HOL
is now fully transparent and, as we think, ready for wider adoption by
philosophers.

On the other hand, there is still room for improvements in
Isabelle/HOL. For example, Isabelle/HOL's Sledgehammer tool, which
orchestrates the calls to external automated theorem provers, such as
\textsc{Leo-II}, still fails to detect the inconsistency of G\"odel's
axioms, while a direct modeling of the problem in TPTP THF syntax in
combination with a direct call of \textsc{Leo-II} succeeds. In other
words, without independent experiments directly in TPTP THF the
inconsistency would thus not have been detected. There are further
technical issues regarding Sledgehammer that motivate
improvements. For example, in default setting, Sledgehammer does not
immediately inform the user when a proof has been found and instead silently
 first executes a series of time-consuming proof analysis processes (e.g. it's
dependency minimizer), before it eventually informs the user of success. For the
involved proof problems presented here (e.g. G\"odel's theorem T3:
\textit{Necessarily, there exists God}), this phase of silence may
take several minutes (during which the user might actually give up on
the proof attempt), while \textsc{Leo-II} already reported success to
Sledgehammer much earlier (e.g. after 2,5 seconds for T3).

Moreover, our work points to a challenge for proof transformation:
even if \textsc{Leo-II}'s proof can eventually be automatically
verified in Isabelle, without a human-intuitive explanation of the
proof, this will presumably be of limited value for philosophers.  The
challenge is to extract from such a verification process an abstract
level argument, similar to what we have presented in this paper: for
philosophers, the reconstruction of such human-intuitive arguments
from machine found proofs will likely be crucial for accepting any
novel results.

Another open problem that we solved in this paper is a fully automatic
proof of T3 directly from Scott's axioms. Again, this proof was
contributed by \textsc{Leo-II}. However, this has been possible only
after providing a more efficient embedding for HOML S5 in HOL.

Both, the automated detection of the inconsistency in G\"odel's axioms
and the fully automatic proof of T3 from Scott's axioms demonstrates
the potential of our AI technology for philosophy: this technology is
in its current state of development already capable of contributing novel results to
metaphysics and to conduct reasoning steps at granularity-levels
beyond common human capabilities.

@@

\small 
%% The file named.bst is a bibliography style file for BibTeX 0.99c
\bibliographystyle{named}
\bibliography{Bibliography}
